---
title: "M4 - Lab 4 - Data Wrangling"
subtitle: "Environmental Data Exploration"
author: "John Fay & Luana Lima"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Answer questions on M3/A3
2. Practice building readable wrangling pipelines
3. Reshape data with `pivot_longer()` and `pivot_wider()`

## Set up your session

Today we will work with the same dataset from our lesson, the North Temperate Lakes Long-Term Ecological Research Station. The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-ntl.71.5), running from 1984-2016. 

```{r workspace setup, message = FALSE}
#Install packages
library(tidyverse)
library(here)

#Read in the data
NTL.phys.data <- read.csv(
  file=here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
)

#Show the datatype of the 'sampledate' column
class(NTL.phys.data$sampledate)
glimpse(NTL.phys.data$sampledate)

# Date conversion 
NTL.phys.data$sampledate <- mdy(NTL.phys.data$sampledate)
```


## Filter

Filtering allows us to choose certain rows (observations) in our dataset.
-   The 1st parameter in the filter function is the dataframe we wish to filter. 
-   The 2nd on is the **filter expression**:
   -    `depth == 0` keeps rows with depth equal to zero (surface)
   -    `lakename %in% c("Paul Lake", "Peter Lake")` keeps Paul & Peter lake rows
   -    `daynum %in% c(152:304)` keeps rows with `daynum` values between 152 and 304
   
```{r filter data}
# note the data types of these two columns
class(NTL.phys.data$lakeid)
class(NTL.phys.data$depth)

# dplyr filtering 
NTL.phys.data.surface <- filter(NTL.phys.data, depth == 0) 

# Choose multiple conditions to filter
summary(NTL.phys.data$lakename)
NTL.phys.data.PeterPaul <- 
  filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake"))

# Choose a range of conditions of a numeric or integer variable
summary(NTL.phys.data$daynum)
NTL.phys.data.JunethruOctober <- filter(NTL.phys.data, daynum %in% c(152:304))

```

```{r Exercise 1 & 2}
# Exercise 1: 
# filter NTL.phys.data for the year 1999
# what code do you need to use, based on the class of the variable?


# Exercise 2: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.


```

Question: Why don't we filter using row numbers?

> Answer: 

---

## Pipes

Pipe is another method to wrangle datasets that looks cleaner and is easier to read.  We designate a pipe with `%>%`. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit. 


```{r Exercise 3 & 4}
#Example using pipes to wrangle data: 
#Add pipes in the correct place below
NTL.phys.data.processed <- 
  NTL.phys.data
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") 
  select(lakename, sampledate:temperature_C) 
  mutate(temperature_F = (temperature_C*9/5) + 32)

#Exercise 3: Using a pipe filter NTL.phys.data for Tuesday Lake from 1990 
# through 1999 and observations collected during July. 

  

#Exercise 4: Using the data from part 3, a pipe, and the summarize() function, 
# find the mean surface water temperature. 
# (hint: in this dataset, surface measurements have depth == 0). 

  
```

Note: There is no difference between summarize() and summarise() in `dplyr`. They are aliases and perform exactly the same operation. Both are maintained and valid.

```{r}
#Exercise 5: Using the NTL.phys.data.processed find the mean temperature for 
#Peter and Paul lake separately.


```

The argument `.groups = "drop"` is not strictly necessary in this particular workflow. It is mainly good practice in case you perform additional operations on the summary data.

## Reshape Data

Use `pivot_longer()` to make a dataset longer and `pivot_wider()` to make it wider.

```{r Exercise 6 & 7}
#Exercise 5: Pivot irradiance data (measured in the water column and measured
#  on the deck of the sampling boat) into one column using pivot_longer. Name
#  the new column holding the irradiance type as "Irradiance_Type", and name the
#  new column holding the irradiance values as "Irradiance_Value".



#Exercise 6: Pivot temperature_C into separate columns based on depth. First
# select the identifying variables needed to make each row unique, i.e, you may 
# want to remove other measurements that are also dependent on depth.



```
