---
title: "M5 - Lab 5 - Data Visualization"
subtitle: "Environmental Data Exploration"
author: "John Fay & Luana Lima"
output: pdf_document
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## Setup

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(here)
library(RColorBrewer)
library(ggthemes)

```

## Import datasets

```{r import-data, message = FALSE, warning = FALSE}
PeterPaul.chem.nutrients <- read.csv(
  here("Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
  ) %>% ymd(sampledate)

EPAair <- read.csv(
  here("Data/Processed_KEY/EPAair_O3_PM25_NC2425_Processed.csv")
  ) %>% ymd(Date)

Neonics <- Neonics <- read.table(
  here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.txt"),
  header = TRUE,
  sep = "|",
  quote = "\"",   
  comment.char = ""
  )

```

---

## Themes: controlling the non-data elements of a plot

A theme controls the non-data elements of a figure. It does **not** change the
values, variables, or geometry being displayed. Themes affect elements such as:

- plot and panel backgrounds;
- titles, subtitles, captions, and axis text;
- grid lines and axis ticks;
- legend appearance and position;
- font family, size, color, and alignment.

Complete themes, such as `theme_minimal()` or `theme_classic()`, establish an
overall design. The `theme()` function then modifies individual elements. Add a
complete theme before `theme()` so later, specific settings are not overwritten.

```{r themes}
surface_temp_1999_plot <- PeterPaul.chem.nutrients %>%
  filter(depth == 0, year4 == 1999) %>%
  ggplot(aes(x = sampledate, y = temperature_C, color = lakename)) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Surface-water temperature in 1999",
    subtitle = "Peter and Paul lakes",
    x = NULL,
    y = "Temperature (C)",
    color = "Lake"
  )


surface_temp_1999_plot + theme_gray()
surface_temp_1999_plot + theme_minimal()

surface_temp_1999_plot +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

When editing a theme, ask whether each element helps the reader. Major grid
lines can support comparisons with the axes; minor grid lines may add clutter.
A legend below a wide plot may work well, whereas direct labels may be better
when only a few series are shown.

```{r useful theme creation template}
my_theme <- theme_base()  + 
  theme(
    #line =              element_line(),
    #rect =              element_rect(),
    #text =              element_text(),
    
    # Modified inheritance structure of text element
    #plot.title =        element_text(),
    #axis.title.x =      element_blank(),
    #axis.title.y =      element_blank(),
    #axis.text =         element_text(),
    
    # Modified inheritance structure of line element
    #axis.ticks =        element_line(),
    #panel.grid.major =  element_line(),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    #plot.background =   element_rect(),
    #panel.background =  element_rect(),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    #legend.position = 'right',
    
    #complete = TRUE
    )
```


```{r reusable-theme}
course_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey30"),
    axis.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

surface_temp_1999_plot + course_theme
```


**Discuss:** Which theme elements make the plot easier to read? Which changes are primarily stylistic?

Resources:

- <https://ggplot2.tidyverse.org/reference/theme.html>
- <https://jrnold.github.io/ggthemes/>

## Color scales and palettes

Color should reflect the type of variable mapped to it.

| Data represented by color | Appropriate palette | Example |
|:--|:--|:--|
| Categories with no inherent order | Qualitative | Lake, site, treatment |
| Values from low to high | Sequential | Temperature, concentration, depth |
| Values extending in two directions from a midpoint | Diverging | Change from baseline, anomaly, residual |

Avoid using a continuous gradient for unrelated categories: the gradient
suggests an order that does not exist. A rainbow palette also introduces uneven
visual emphasis and is often difficult for readers with color-vision
deficiencies.

### ColorBrewer

The `RColorBrewer` package provides qualitative, sequential, and diverging
palettes designed for maps and other data visualizations.

```{r colorbrewer-palettes}
display.brewer.all(type = "qual")
display.brewer.all(type = "seq")
display.brewer.all(type = "div")

surface_temp_1999_plot +
  scale_color_brewer(palette = "Dark2") +
  course_theme
```

For a discrete variable, use `scale_color_brewer()` or
`scale_fill_brewer()`. Check the number of colors supported by a palette before
using it for a variable with many categories.

### Viridis

Viridis palettes are perceptually uniform: equal changes in the data produce
approximately equal perceived changes in color. They retain useful contrast
when printed in grayscale and are designed to work better for common forms of
color-vision deficiency.

```{r viridis-example}
PeterPaul.chem.nutrients %>%
  filter(!is.na(temperature_C), !is.na(depth)) %>%
  ggplot(aes(x = sampledate, y = temperature_C, color = depth)) +
  geom_point(size = 0.7, alpha = 0.7) +
  scale_color_viridis_c(
    option = "magma",
    direction = -1,
    name = "Depth"
  ) +
  labs(x = NULL, y = "Temperature (degrees C)") +
  course_theme
```

The `_c` suffix indicates a continuous scale; `_d` indicates a discrete scale.
Changing `direction` from `1` to `-1` reverses the palette. The direction should
support the meaning of the variable—for example, darker colors might
intuitively represent greater depth.

## Designing accessible plots

Accessibility is not achieved by selecting one particular palette. Use several
strategies together:

1. Do not rely on color alone. When feasible, combine color with shape,
   linetype, facets, or direct labels.
2. Use sufficient contrast between marks, text, and the background.
3. Use readable text and export the plot at an appropriate size.
4. Limit the number of colors and use palettes designed for the variable type.
5. Label variables and units clearly; do not make the reader decode column names.
6. Check that categories remain distinguishable in grayscale and under common
   color-vision-deficiency simulations.
7. Use transparency carefully: it can reveal overlapping points, but very low
   opacity can reduce contrast.

Here, lake is represented by both color and linetype, so the distinction does
not depend only on color.

```{r redundant-cues}
surface_temp_1999_plot +
  aes(linetype = lakename) +
  scale_color_brewer(palette = "Dark2") +
  labs(linetype = "Lake") +
  course_theme
```

**Discuss:** When might redundant cues improve a graph, and when might they make
it too busy?

---

## Exercise 1: Temperature across years and lakes

Use `PeterPaul.chem.nutrients` to plot surface-water temperature by day of year.

**Plot requirements**

1. Filter to surface observations (`depth == 0`) with non-missing temperatures.
2. Map day of year to the x-axis and temperature to the y-axis.
3. Map year to point color. Decide whether year should be a numeric gradient or
   a categorical variable and choose a suitable scale.
4. Use point size and transparency that make overlap visible.
5. Facet by lake in two rows.
6. Add an informative title, axis labels including units, and a clear color
   legend title.
7. Apply a readable theme and remove elements that do not help the comparison.

```{r exercise-1, fig.height = 4.5, fig.width = 7, warning = FALSE}
# Build the plot here.

```

**Interpretation**

1. What seasonal pattern is visible in surface-water temperature?
2. Do the timing or magnitude of peak temperatures differ between lakes?
3. Is year easier to interpret as a continuous gradient or as separate
   categories in this plot? Explain your choice.
4. Where does overplotting occur, and how did your formatting address it?

## Exercise 2: Temperature over time and depth

Plot temperature across the full sampling record and use color for depth.

**Plot requirements**

1. Map sampling date to the x-axis, temperature to the y-axis, and depth to
   point color.
2. Use points with `size = 0.5` and select a suitable transparency.
3. Apply the continuous viridis `magma` palette. Compare `direction = 1` and
   `direction = -1`, then retain the direction that best communicates depth.
4. Format the date axis with labeled breaks every five years.
5. Use facets or another deliberate strategy to distinguish the lakes.
6. Add a descriptive title, axis labels with units, and a useful legend title.
7. Preserve useful reference lines without making the dense series harder to read.

```{r exercise-2, fig.height = 4.5, fig.width = 7, warning = FALSE}
# Build the plot here.

```

**Interpretation**

1. How does temperature vary with depth?
2. Does the relationship appear constant across seasons and years?
3. Which palette direction did you choose, and what does the dark end convey?
4. What is clarified—or hidden—by the way you separated the lakes?

## Exercise 3: Relationship between ozone and PM2.5 AQI

Use `EPAair` to examine the relationship between ozone and PM2.5 AQI values.

**Plot requirements**

1. Map ozone AQI to the x-axis and PM2.5 AQI to the y-axis.
2. Map latitude to point color using an appropriate continuous palette.
3. Make points 50% transparent and choose a size that keeps dense areas visible.
4. Add a linear line of best fit. Decide whether its confidence interval helps.
5. Use reader-friendly axis labels rather than original column names.
6. Add a title and a legend title that describes latitude clearly.
7. Apply formatting that focuses attention on the relationship.

```{r exercise-3, fig.height = 4.5, fig.width = 6, warning = FALSE}
# Build the plot here.

```

**Interpretation**

1. Is the association positive, negative, or weak? What supports your answer?
2. Are there clusters, outliers, or changes in spread that the fitted line hides?
3. Does latitude reveal a meaningful spatial pattern?
4. What does the fitted line summarize, and what can it **not** establish about
   causation?

## Exercise 4: Comparing PM2.5 AQI distributions by year

Create and compare several plots of PM2.5 AQI grouped by year—for example, a
boxplot, violin plot, ridge plot, or points combined with a summary.

**Plot requirements**

1. Treat year as a categorical variable.
2. Create at least two reasonable plot types before selecting one.
3. Use the same axis limits, labels, and ordering when comparing alternatives.
4. Select a qualitative palette, or one neutral fill if color adds no information.
5. Include an explanation of the AQI measure in the axis label.
6. Make titles, text, and grid lines consistent across alternatives.
7. Present the plot you think communicates the distribution most effectively.

```{r exercise-4, fig.height = 4.5, fig.width = 6, warning = FALSE}
# Create and compare candidate plots here.

```

**Interpretation**

1. How do center, spread, skew, and potential outliers differ by year?
2. Which distributional features are visible in your chosen plot type?
3. What would a reader lose if you displayed only the mean for each year?
4. Why is your selected plot more effective than the alternative you tested?

## Exercise 5: Counts and factor order

The graph below displays counts of endpoints measured in neonicotinoid ecotoxicology studies, but the default graph makes comparison difficult. 

```{r}
# Original plot
ggplot(Neonics) +
  geom_bar(aes(x = Endpoint.))
```

Improve it without changing the underlying data.

- Use a single fill color unless color encodes additional information.
- Put the most common endpoint at the top.
- Use readable labels. You may need to orient the bars so endpoint names are easy to read.
- State why `geom_bar()` or `count()` plus `geom_col()` is appropriate.

```{r ordered-bars}
# Improve the plot here.
```

> **Why did you choose this bar-chart workflow?**


**Interpretation**

1. Which endpoints appear most and least frequently?
2. Why does ordering by frequency make the plot easier to interpret?
3. Does color encode information in your revision? If not, why is a consistent
   fill preferable?
4. What is one conclusion supported by this figure, and what is one question it
   cannot answer?

## Exercise 6: Zoom without discarding data

Create a scatterplot of `irradianceWater` and `dissolvedOxygen`. Zoom to x = 0–250 and y = 0–20 using `coord_cartesian()`.

Then replace it temporarily with `xlim()` and `ylim()`. Record any warning and explain why the two methods are not analytically equivalent, especially when a smoother is included.

```{r ordered-bars}
# Add your code here
```

## Final check

Before submitting or sharing any figure, ask:

- Can a reader understand the main comparison without reading the code?
- Are variables, units, and categories clearly labeled?
- Is any distinction based on color alone?
- Is the palette appropriate for the data type?
- Are the text and symbols readable at the exported size?
- Does every theme element support the message, or can something be removed?
