---
title: "Assignment 3: Data Exploration"
author: "Henry Park"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Exploration M3: Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2016-26_raw.csv). Name these datasets "Neonics" and "Litter", respectively.

```{r setup,  message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse);
library(lubridate);
library(here)

Neonics <- here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.txt")
Litter <- here("Data/Raw/NEON_NIWO_Litter_massdata_2016-26_raw.csv")
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: One possible reason for investigating this link might be to examine what impact neonicotinoids have on different insect populations.There are many insect species which perform important natural functions in the world, and understanding if a certain class of insecticide might potentially threaten these insect species is valuable information to determmine if neonicotinoids are a more useful or a more harmful chemical to use. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: After searching up on the subject, I learned that forest litter and debris performs some important environmental functions. Notably, two aspects of it I had learned about was that forest litter creates an environment suitable for certain organisms, such as other plants, fungi, lichens, and insects. Forest litter also contributes to the humidity that can sustain vegetation during dry spells. It might therefore be important to monitor the levels of forest litter to see the overall health of an ecosystem. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Each sampling site used a mix of an elevated trap and a ground-level trap for collection. 
 2.Samples were collected at different time intervals for both trap categories.
 3.For sampling purposes, debris were divided into 'functional groups' such as leaves, needles, twigs, etc.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
Neonics_data <- read.csv(
  file=here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.txt"), sep = "|"
  )

#The dataset's dimensions is 8478 rows/objects by 76 columns/variables.
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}
Effects <- factor(Neonics_data$Effect)
summary(Effects)
```

> Answer: The most common effect which was studied was Mortality, totaling at 2,687 occurrences, followed by Population at 1,377 occurrences. These two effects might be useful to study to consider how insecticides are affecting insect populations, observing mortality rates and seeing the ebbs and flows of the general population and which groups of the population are more or less resistant to the insecticide.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}
Common_Names <- factor(Neonics_data$X.Species.Common.Name)
summary(Common_Names, maxsum=7)
```

> Answer: The six most commonly-studied species are: Honey Bees, Carniolan Honey Bees, Parasitic Wasps, Bumble Bees, Italian Honeybees, and Asiatic Honey Bees. Most of these species are bee species, and all in all they are all hymenoptera species. Notably, bee species are well-established pollinators who interact often with agriculture and play a large role in it, so they are a critical species to maintain stable populations especially when considering an insecticide that might endanger insect populations. 

8.  Concentrations are always a numeric value. What is the class of `Conc.1.Mean..Author..` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
class(Neonics_data$X.Conc.1.Mean..Author..)
```

> Answer: The class for 'Conc.1.Mean..Author..' column is character-based. Looking at the values in the dataframe itself, there are a lot of other symbols in the values beyond just the number digits themselves, and R may not recognize the other symbols and thus be forced to classify the column as a character-based one. 

## Explore your data graphically (Neonics)

9. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}
ggplot(Neonics_data, aes(x=Endpoint.)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

> Answer: The two most common end points are LOEL and NOEL. LOEL is defined as Lowest Observed Effect Level and NOEL is defined as No Observed Effect Level. 

## Explore your data (Litter)

10. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}
Litter_data <- read.csv(
  file=here("Data/Raw/NEON_NIWO_Litter_massdata_2016-26_raw.csv")
  )

class(Litter_data$collectDate)
Litter_data$collectDate <- ymd(Litter_data$collectDate)

unique(Litter_data$collectDate)
```

> Answer: The dates litter was sampled in August 2018 was: 2018-08-02, 2018-08-30.

11. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}
unique(Litter_data$plotID)
```

> Answer: The different IDS include: "NIWO_047" "NIWO_061" "NIWO_062" "NIWO_057" "NIWO_058" "NIWO_064" "NIWO_051" "NIWO_067" "NIWO_063" "NIWO_041" "NIWO_046" "NIWO_040". The unique function shows off all of the different values once no matter how many instances are in the original dataset, and summary provides summary statistics and the frequency that each unique value appeared.

12. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}
ggplot(Litter_data, aes(x=functionalGroup)) +
  geom_bar()

```


What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles appear to have the highest biomass at these sites. 

13. Write a short data-exploration summary for Neonics.

> - **One key characteristic:** Conducting 8748 insecticide tests on various insect populations, observing characteristics of certain effects such as mortality rates and population changes.
> - **One data-quality issue:** Certain elements of the data are not classified properly, such as 'X.Conc.1.Mean..Author..' introducing unknown characters into a numeric column and forcing it to be classified as a character column. 
> - **One interesting pattern:** Plotting out the Endpoints shows most the test studies measuring LOEL and NOEL, measuring the lowest possible point where an insecticide causes an adverse effect or measuring levels where no visible effect is noticeable. 
> - **One question for further analysis:** Which insecticides cause adverse effects along a select number of insect species, and which are particularly lethal across all insect species? 

14. Write a short data-exploration summary for Litter. 

> - **One key characteristic:** Conducting 4443 checking tests across various locations to observe quantities of wood litter, categorized by litter class and biomass.
> - **One data-quality issue:** Certain data columns were not yet transformed into a usable format for analysis, such as collectDate classified as general character-based data and not specific Dates. 
> - **One interesting pattern:** Biomass totals for various types of wood litter were relatively even across the board, but needles skewed with the largest amount of biomass across all tests. 
> - **One question for further analysis:** How do biomass results change over time as ecosystems change and are affected by climate change in future years of analysis? Are there trends of litter totals moving up or down? 
