---
title: "Assignment 5: Data Visualization"
author: "Henry Park"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This assignment accompanies the Environmental Data Analytics module on data
visualization. You will create, format, combine, and interpret plots using
environmental datasets. Your goal is not only to produce correct code, but also
to make deliberate choices about themes, color, accessibility, and the message
communicated by each figure.

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.
7. All figures should include clear labels and units, readable text, and color
   choices appropriate for the type of variable displayed. Do not rely on color
   alone when another cue (such as shape, linetype, or facets) would make the
   figure more accessible.
8. Complete your own code before using AI in the designated comparison activity
   in Question 7. Include the prompt and your evaluation of the AI response.

---

## Set up your session 

1. Set up your session. Load the `tidyverse`, `here`, `cowplot`, and any color
palette package(s) you choose to use. Verify your project directory. Read in the
processed NTL-LTER nutrients and chemistry/physics file for Peter and Paul Lakes
(`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`) and the processed
Niwot Ridge litter file (`NEON_NIWO_Litter_mass_trap_Processed.csv`) from the
`Processed_KEY` folder.

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse);
library(here);
library(cowplot);

getwd()

PeterPaul <- read.csv(file=here('Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv'))
LitterMass <- read.csv(file=here('Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv'))

#2 
PeterPaul$sampledate <- ymd(PeterPaul$sampledate)
LitterMass$collectDate <- ymd(LitterMass$collectDate)
class(PeterPaul$sampledate)
class(LitterMass$collectDate)
```

## Define your theme
3. Build a theme and set it as your default theme. Start with a complete theme,
such as `theme_minimal()` or `theme_classic()`, and then customize at least two
of the following:
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

Choose a base font size that will remain readable when the plots are knitted.
Briefly identify the two or more elements you changed and explain how the
changes improve readability.

```{r build plot theme}
#3
assignment_theme <- theme_minimal(base_size = 12) +
  theme(
    text = element_text(size = 12),
    panel.background = element_rect(color = "lightblue"),
    plot.title = (element_text(face = "bold",
                               size = 16,
                               color = "black")),
    legend.position = "right",
    axis.title = element_text(face = "bold",
                              size = 14),
    axis.text = element_text(size = 11,
                             color = "black")
  )
```

> Theme choices and rationale: I chose the minimal theme as a base and adjusted text sizes for the large title and axis titles to make them more distinguishable and visible. I also slightly adjusted the elements on the axes themselves to be slightly bigger, and added a blue border around the actual plots to make them less intrusive on the graph. I mostly aimed to make a theme where the plot points themselves and the main titles popped out before any of the other graphical elements. 

## Create graphs
For Questions 4–7, create `ggplot2` graphs and adjust their aesthetics using
data-visualization best practices. Match the color palette to the variable type,
label variables and units clearly, and remove formatting that does not support
the comparison.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), using
separate aesthetics for Peter and Paul lakes. Do not rely only on color: also
use shape, linetype, facets, or another appropriate visual cue. Add a separate
line of best fit for each lake using the `lm` method. Zoom in so that the central
pattern is visible without deleting the extreme observations from the analysis
(hint: use `coord_cartesian()` rather than `xlim()` or `ylim()`). Include a
descriptive title, axis labels with units, and a meaningful legend title.

```{r plot total P vs PO4}
#4 
Q4 <- 
  PeterPaul %>%
  ggplot(aes(x = po4, y = tp_ug, linetype = factor(lakename))) +
  geom_point(alpha = 0.5) +
  facet_wrap(vars(lakename), nrow = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Plot of Phosphorus by Phosphate in Peter and Paul Lakes",
       x = "Phosphate (po4)",
       y = "Phosphorus (tp_ug)",
       linetype = "tp_ug increase by po4 increase") +
  coord_cartesian(xlim = c(0,50),
                  ylim = c(0,150)) +
  assignment_theme

Q4
```

Question: What relationship do you observe between phosphate and total
phosphorus? Does the relationship appear to differ between lakes? Identify any
extreme observations and explain why zooming with `coord_cartesian()` is
preferable to removing them with scale limits.

> Answer: For both lakes, the two appear to linearly increase with one another--when phosphate increases, total phosphorus increases as well. The relationship does slightly differ between lakes, as total phosphorus by phosphate rises more sharply in Peter Lake compared to Fall Lake, but is not significantly different in behavior. Zooming in with coord_cartesian is important as we can likely see the impact of an outlier value on the regression line for Paul Lake--showing that these outlier values do have a real impact on the conclusions we get from the data. Removing them outright may skew the data and make it inaccurate. 


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c)
TN, with month on the x-axis and lake represented by a consistent aesthetic.
Use a qualitative, color-vision-friendly palette. Then create a `cowplot` that
combines the three graphs. Make sure that only one legend is present, the month
order is correct, graph axes are aligned, and each y-axis states the variable
and units.

Tips:
* Month should appear in calendar, not alphabetical, order. You will need to convert 
to factor to order it.
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
PeterPaul <- PeterPaul %>%
  mutate(Month = month(sampledate))

#5 
Q5Temp <-
  PeterPaul %>%
  ggplot(aes(x = Month, y = temperature_C, color = lakename)) +
  geom_boxplot() +
  labs(color = element_blank(),
       x = "Month",
       y = "Temperature (degrees Celsius)") +
  assignment_theme +
  theme(legend.position = "none")
Q5Temp

Q5TP <-
  PeterPaul %>%
  ggplot(aes(x = Month, y = tp_ug, color = lakename)) +
  geom_boxplot() +
  labs(color = element_blank(),
       x = "Month",
       y = "Total Phosphorus (tp_ug)") +
  assignment_theme +
  theme(legend.position = "none")
Q5TP

Q5TN <-
  PeterPaul %>%
  ggplot(aes(x = Month, y = tn_ug, color = lakename)) +
  geom_boxplot() +
  labs(color = element_blank(),
       x = "Month",
       y = "Total Nitrogen (tn_ug)") +
  assignment_theme +
  theme(legend.position = "none")
Q5TN

CombinedLegend <- get_legend(Q5Temp + assignment_theme)

Q5Combined <- plot_grid(Q5Temp, Q5TP, Q5TN, 
                        labels = c("Temperature", "TP", "TN"), 
                        CombinedLegend, ncol = 4)
Q5Combined
```

Question: What do you observe about the center, spread, and potential outliers
of each variable across seasons and between lakes? Support your interpretation
with specific features visible in the boxplots.

> Answer: Paul Lake appears to have slightly higher temperatures, while Peter Lake appears to have higher total amounts of phosphorus and nitrogen in the water. Comparing the phosphorus and nitrogen values, Paul Lake's values appear more bunched up in line with one another, while Peter Lake has more outlier values in comparison. However, the boxplots themselves indicate that taken in aggregate, the median values between the two lakes are not far off from one another. 

6. [Niwot Ridge] Subset the litter dataset to the `"Needles"` functional group.
Plot dry mass of needle litter by date and distinguish NLCD classes using a
qualitative color aesthetic. You do not need to rename the land-use classes.
Format the date axis, labels, legend, point size, and transparency so that
overlapping observations remain visible.

7. [Niwot Ridge] First, independently create a second version of the plot with
the three NLCD classes separated into facets rather than distinguished by color.
Keep other formatting choices as consistent as possible so that you can compare
the effect of faceting itself.

```{r Plot litter}
#6
Needle_Litter <-
  LitterMass %>%
  filter(functionalGroup == 'Needles')

Needle_Graph <-
  Needle_Litter %>%
  ggplot(aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_point(alpha = 0.7) +
  labs(title = "Needle Litter Dry Mass by NLCD Classes",
       color = "NLCD Classes",
       x = "Collection Dates",
       y = "Dry Mass") +
  assignment_theme
Needle_Graph

#7
Needle_Facet <-
  Needle_Litter %>%
  ggplot(aes(x = collectDate, y = dryMass)) +
  geom_point(alpha = 0.7) +
  facet_wrap(vars(nlcdClass)) +
  labs(title = "Needle Litter Dry Mass by NLCD Classes",
       color = "NLCD Classes",
       x = "Collection Dates",
       y = "Dry Mass") +
  assignment_theme
Needle_Facet
```

## AI comparison

8. After completing your own faceted plot, provide an AI tool with the Question 7 instructions and the relevant variable names. Ask it to propose `ggplot2` code for the same figure. Do **not** replace your original solution with the AI version.

Paste the exact prompt you used below:

> AI tool used: Google

> AI prompt: Using the Needle_Litter dataframe, make a ggplot titled "Dry Mass of Needle Litter Over Time" graphing dryMass by collectDate and distinguishing nlcdClasses separated into facets.

Run and inspect the AI-generated code, correcting only errors required to make
it run. If you make corrections, identify them. Then compare the two solutions.

```{r AI solution}
# Paste the AI-generated code here. Clearly comment any corrections you make.

Needle_AI <-
  Needle_Litter %>%
  ggplot(aes(x = collectDate, y = dryMass)) +
  geom_point(alpha = 0.6) + 
  facet_wrap(~ nlcdClass, scales = "free_y") +
  theme_minimal() +                             
  labs(
    title = "Dry Mass of Needle Litter Over Time",
    x = "Date",
    y = "Dry Mass (g)",
    color = "NLCD Class"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
Needle_AI

#Corrections: On the ggplot function, there was an incorrect argument with a 
#'data' function.
#Corrections: On initial creation, the graph connected all of the dots in the 
#plot. That function was removed.
#Corrections: Color was initially added to the geom_point line. 

```

9. Which plot—your solution or the AI solution—more effectively communicates the data? Evaluate at least three specific choices, such as data filtering, facet structure, date-axis formatting, labels and units, theme, accessibility, or unnecessary code. Identify one useful idea from the AI
solution and one choice you would reject or improve. If the solutions are very similar, explain why both approaches are appropriate.

> AI comparison: While there are not any particular differences in the general 'functionality' of the graph--that is, just the ability to see the spread of data throughout the plot--the theming we set for the assignment prior to starting made the manually-created one easier to read and interpret. The AI did attempt to include colors along with the facet map, which could be a nice visual touch (while not strictly necessary because the points are already faceted, it would just help further distinguish the different land classes). For the purposes of making the graph in this assignment, both were adequately fine barring some slight manual tuning of the AI-generated one. 

10. Question: Which encoding in Questions 6 and 7—color or facets—do you think is more effective for comparing NLCD classes, and why? Consider overlap, ease of comparison, accessibility, and the amount of space used by each figure.

> Answer: When it came to my results, I found that the faceted graphs felt easier to interpret and compare. I think the spacing out of all of the results made finding patterns easier to see-that being said, I can also see a strong benefit in placing all three categories into the same graph distinguished by color so that I can start plotting out patterns all on the same plot. For avoiding overlap, I much preferred the faceted graph--but I can see the value of the color graph for being more economical with space and making it easier to directly compare patterns. 
